Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-03-23 bbbq_article.tex
Made float smaller to remove warning
Feedback @fransbianchi

% \paragraph{Data sets for TMH epitopes}

We used a human, viral and bacterial reference proteome to 
determine the percentages of epitopes overlapping
with TMHs.
We used a human reference proteome
with UniProt ID UP000005640\_9606, which
is a representative set of the full proteome
and used is our previous study (\cite{bianchi2017}).
We used SARS-CoV-2 as our viral reference proteome
with UniProt ID UP000464024 and
Mycobacterium tuberculosis (MTb) 
with UniProt ID UP000001584.
We chose these pathogens based on their impact on the human population.

@@

To determine if epitopes derived from TMHs are presented at all,
we started from epitopes identified in elution studies
for MHC-I (\cite{schellens2015comprehensive}) 
and MHC-II (\cite{bergseng2015different}).
For each of these peptides, the human representative reference proteome
was searched for its origin.
We kept only the epitopes which were uniquely present
in the proteome.
We predicted the topology of the proteome
and counted how often the epitopes overlapped
with a predicted TMH.
In this analysis, both TMHMM and 
PureseqTM (see section 'Prediction software used') 
were used to predict the topology.
The full analysis can be found
at \url{https://github.com/richelbilderbeek/bbbq_article_issue_157}.

@@

To determine the percentages of MHC-I and MHC-II epitopes overlapping
with TMHs, we used mostly the same analysis as described in \cite{bianchi2017}.
To summarize: from a proteome all possible 9-mers are derived. For each
of these peptides, it was determined if it was part of a 
TMH (at least overlapping with 1 residue), 
and if the peptide bound to an MHC-I haplotype.
For MHC-II, 14-mers were used, as these are the most frequently occurring
epitope length (\cite{bergseng2015different}).

This study differs in some aspects from our previous work (\cite{bianchi2017}), 
described below in more detail.
The main differences are our definition of what a so-called binder is, 
the inclusion of both MHC-I and MHC-II haplotypes, 
the prediction software used, 
and the significance level to determine if TMH-derived peptides will bind.
These deviations are either a refinement of our previous approach or
a pragmatic choice made due to the extension of the original experiment.
Additionally, instead of only using a human proteome, this study
also includes a viral and bacterial proteome.

The definition of a binder differs from \cite{bianchi2017}:
in our current study a peptide was called a binder if, within a certain haplotype, 
any of its 9-mer peptides have an IC50 value in the lowest 2\% of 
the epitopes within a 
\emph{proteome} (see tables \ref{tab:ic50_binders_mhc1} and \ref{tab:ic50_binders_mhc2}
for values), whereas the original study defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17
We believe our revised definition precludes bias of proteins 
that give rise to no or only very few MHC epitopes

The 13 MHC-I haplotypes used in this study are the same as 
we used in our previous study (\cite{bianchi2017}).
The MHC-II haplotypes used additionally were selected 
to occur with a phenotypic frequency of at least 14\% in
the human population (\cite{greenbaum2011functional}),
resulting in 21 haplotypes.
When using an MHC-II haplotype, instead of using 9-mers, 14-mers were
used, as these are the most common MHC-II epitope size,
as found in the elution study (\cite{bergseng2015different}).

@@

To detect the evolutionary conservation of TMHs, 
we collected human mutations and 
tallied their predicted location.
To be more precise, we collected single nucleotide
polymorphisms (SNPs) within the human population
that resulted in the substitution of one amino acid.
We then predicted that protein's topology and tallied if
the SNP occurred within a TMH or not, 
after which we used statistics to determine if there are
more, less or an equal amount of SNPs in TMHs
as a measure of evolutionary conservation.
This workflow is discussed in more detail below.

@@

As a data source, multiple
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases were used,
which are 'gene', to find the gene names of membrane proteins, 
'dbSNP' (\cite{sherry2001dbsnp}) for SNPs associated with those genes
and 'protein', to obtain the sequence of proteins that SNPs act upon.
The 'dbSNP' contains 650 million 
catalogued non-redundant humane variations (called RefSNPs,
\url{https://www.ncbi.nlm.nih.gov/snp/docs/RefSNP_about/}).

@@

To retrieve the data from these databases the
\verb;rentrez; R package (\cite{rentrez}) was used
that calls the NCBI website's API. To provide for a 
stable user experience for all users, 
this API limits the user to 3 calls per second.
Additionally, the API splits the result of a bigger
query into multiple pages, each of which needs one API call.
We wrote the \verb;sprentrez; package (\cite{sprentrez}) to provide for 
bigger queries of multiple (and delayed) API calls.

@@

The number of variations
was limited to the first 250 variations per gene (see figure
\ref{fig:snps_per_gene_name_ncbi} for the number of SNPs
within the dbSNP database for these gene names),
resulting in 60,683 variations at the protein level.
A variation needs not always be a SNP,
as dbSNP also catalogs other DNA alterations, such as, among others, insertions,
deletions and tandem repeats.
We select only the variations that result in a SNP for
a single amino acid substitution, resulting in 38,882 SNPs.

% \paragraph{Selection of SNPs}
%
We selected to analyze SNPs with the lowest SNP ID first.
This means we have a bias for picking SNPs with
an earlier discovery date.
We expect this bias to be biologically irrelevant.
To verify this expectation, we determined the
relative position of each SNP in a protein,
as shown in figure \ref{fig:snp_rel_pos}.

@@

\begin{table}[]
  \begin{tabular}{llll}
    Goal & Tool & Reference \\ 
    \hline
    Predict topology                  & TMHMM                     & \cite{krogh2001predicting} \\
    Predict topology                  & PureseqTM                 & \cite{wang2019efficient} \\
    Predict epitopes MHC-I            & \verb;epitope-prediction; & \cite{bianchi2017} \\
    Predict epitopes MHC-II           & NetMHCIIpan               & \cite{nielsen2008quantitative,karosiene2013netmhciipan} \\
    Call TMHMM from R                 & tmhmm                     & \cite{tmhmm} \\
    Call PureseqTM from R             & pureseqtmr                & \cite{pureseqtmr} \\
    Call NetMHCIIpan from R           & netmhc2pan                & \cite{netmhc2pan} \\
    Combine all                       & bbbq                      & \cite{bbbq}
  \end{tabular}
  \caption{
    Overview of all software used in this research.
  }
  \label{table:software_used}
\end{table}

@@

\begin{figure}[!htbp]
  \includegraphics[height=0.9\textheight]{bbbq_1_smart_results/fig_f_tmh_mhc2_2_grid.png}
  \caption{
    Percentage of MHC-II epitopes overlapping with TMHs
    for a human, viral and bacterial proteome.
    The horizontal lines indicate the percentage as expected by chance.
    See table \ref{tab:tmh_binders_mhc2} for the exact TMH and epitope counts.
    Note that for smaller proteomes a percentage of zero is likelier.
  }
  \label{fig:2}
\end{figure}

>>>>

% \paragraph{Data sets for TMH epitopes}
A human, viral and bacterial reference proteome was used to determine the percentages of epitopes overlapping with TMHs. The human reference proteome has UniProt ID UP000005640_9606, which is a representative set of the full proteome and also used in our previous study (\cite{bianchi2017}).  SARS-CoV-2 with reference proteome UniProt ID UP000464024 was used as a viral proteome and Mycobacterium tuberculosis (MTb) with UniProt ID UP000001584 was used . These pathogens were chosen based on their impact on the human population(references).

@@

To determine if epitopes derived from TMHs are presented we used epitopes identified in \frans{Welke studies is dit dan van SARS-COV2 en MTB ?}
for MHC-I (\cite{schellens2015comprehensive}) 
and MHC-II (\cite{bergseng2015different})peptide elution studies.
For each of these peptides, the human representative reference proteome was searched for its origin\frans{wat bedoel je?}. We filtered for epitopes that were uniquely present in the proteome and correlated these for overlap with predicted transmembrane regions in the proteome. 
In this analysis, both TMHMM and PureseqTM (see section ’Prediction software used’) were used to predict the transmembrane topology. 
The full analysis can be found
at \url{https://github.com/richelbilderbeek/bbbq_article_issue_157}.

@@

To determine the percentages of MHC-I and MHC-II epitopes overlapping with TMHs, we applied a similar analysis strategy as described in  \cite{bianchi2017}.
To summarize: from a proteome all possible 9-mers are derived. For each
of these peptides, it was determined if it was part of a 
TMH (at least overlapping with 1 residue), 
and if the peptide bound to an MHC-I haplotype.
For MHC-II, 14-mers were used, as these are the most frequently occurring
epitope length (\cite{bergseng2015different}).

This study differs in some aspects from our previous work (\cite{bianchi2017}), 
The definition of a binder differs from \cite{bianchi2017}:
in the current study a peptide is called a binder if, within a certain haplotype, 
any of its 9-mer peptides have an IC50 value in the lowest 2\% of 
the epitopes within a 
\emph{proteome} (see tables \ref{tab:ic50_binders_mhc1} and \ref{tab:ic50_binders_mhc2}
for values), whereas the original study defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17
We believe our revised definition precludes bias of proteins 
that give rise to no or only very few MHC epitopes

The 13 MHC-I haplotypes used in this study are the same as 
used in our previous study (\cite{bianchi2017}).
The MHC-II haplotypes used additionally were selected 
to occur with a phenotypic frequency of at least 14\% in
the human population (\cite{greenbaum2011functional}),
resulting in 21 haplotypes.
When using an MHC-II haplotype, instead of using 9-mers, 14-mers were
used, as these are the most common MHC-II epitope size,
as found in the elution study (\cite{bergseng2015different}).\frans{what percentage is 9 mer or 14 mer?}

@@

To detect the evolutionary conservation of TMHs, 
single nucleotide
polymorphisms (SNPs) within the human population
that resulted in the substitution of one amino acid were collected from ??? and tallied if
the SNP occurred within a TMH or not. Using statistics is was determined if there are more, less or an equal amount of SNPs in TMHs
as a measure of evolutionary conservation.
This workflow is discussed in more detail below.

@@

As a data source, multiple
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases were used,
which are 'gene',\frans{snap ik niet which are gene?} to find the gene names of membrane proteins, 
'dbSNP' (\cite{sherry2001dbsnp}) for SNPs associated with those genes
and 'protein', to obtain the sequence of proteins that SNPs act upon.
The 'dbSNP' contains 650 million 
catalogued non-redundant humane variations (called RefSNPs,
\url{https://www.ncbi.nlm.nih.gov/snp/docs/RefSNP_about/}).

@@

\frans{ dit is relevant waarom: ? To retrieve the data from these databases the
\verb;rentrez; R package (\cite{rentrez}) was used
that calls the NCBI website's API. To provide for a 
stable user experience for all users, 
this API limits the user to 3 calls per second.
Additionally, the API splits the result of a bigger
query into multiple pages, each of which needs one API call.
We wrote the \verb;sprentrez; package (\cite{sprentrez}) to provide for 
bigger queries of multiple (and delayed) API calls.}

@@

The number of variations
was limited to the first 250 variations per gene (see figure
\ref{fig:snps_per_gene_name_ncbi} for the number of SNPs
within the dbSNP database for these gene names),
resulting in 60,683 variations at the protein level.
Only variations that result in a SNP for
a single amino acid substitution were analysed, resulting in 38,882 SNPs.

% \paragraph{Selection of SNPs}
%
SNPs were picked based on ID number, which is linked to their discovery date. To verify that ID number is unrelated to SNP positions the relative positons of the analyzed SNPS in a protein was determined and showed no positional bias, shown in figure \ref{fig:snp_rel_pos}.

@@

[RJCB: I suggest to move this section to the Supplementary Materials]
\frans{lijkt mij een goed idee}
\begin{table}[]
  \begin{tabular}{llll}
    Goal & Tool & Reference \\ 
    \hline
    Predict topology                  & TMHMM                     & \cite{krogh2001predicting} \\
    Predict topology                  & PureseqTM                 & \cite{wang2019efficient} \\
    Predict epitopes MHC-I            & \verb;epitope-prediction; & \cite{bianchi2017} \\
    Predict epitopes MHC-II           & NetMHCIIpan               & \cite{nielsen2008quantitative,karosiene2013netmhciipan} \\
    Call TMHMM from R                 & tmhmm                     & \cite{tmhmm} \\
    Call PureseqTM from R             & pureseqtmr                & \cite{pureseqtmr} \\
    Call NetMHCIIpan from R           & netmhc2pan                & \cite{netmhc2pan} \\
    Combine all                       & bbbq                      & \cite{bbbq}
  \end{tabular}
  \caption{
    Overview of all software used in this research.
  }
  \label{table:software_used}
\end{table}

@@

\begin{figure}[!htbp]
  \includegraphics[height=0.8\textheight]{bbbq_1_smart_results/fig_f_tmh_mhc2_2_grid.png}
  \caption{
    Percentage of MHC-II epitopes overlapping with TMHs
    for a human, viral and bacterial proteome.
    The horizontal lines indicate the percentage as expected by chance.
    See table \ref{tab:tmh_binders_mhc2} for the exact TMH and epitope counts.
    Note that for smaller proteomes a percentage of zero is likelier.
  }
  \label{fig:2}
\end{figure}
